---
title: "Applied Biostatistics 1 Introduction"
author: "CUNY Graduate School of Public Health and Health Policy"
output: ioslides_presentation
---

## Applied Biostatistics I

Review of Simple Linear Regresssion

## Example: blood pressure and aging

Hypothesis: age-related physiological changes (narrowing and hardening) of
blood vessels tend to raise blood pressure.

In order to investigate the hypothesis, what would you do first? (With data on
age and blood pressure.)

## Systolic Blood Pressure by Age

* Using a subsample of NHANES 2009-2010 (in the `NHANES` R package)

```{r,echo=FALSE}
library(NHANES)
data(NHANES)
miniH <- NHANES[NHANES$SurveyYr == "2009_10", c("Age", "BPSysAve")]
miniH <- miniH[miniH$Age >= 18 & complete.cases(miniH), ]
plot(miniH)
```

## Local Averaging

* Mean SBP for the 20s, 30s, 40s, ...

```{r,echo=FALSE}
plot(miniH)
lines(
    gplots::wapply(
        miniH$Age, miniH$BPSysAve, method = "width", width = 10, n = 10),
    col = "red", lwd = 3)
```

## Linear Regression

We fit a straight line to the points on the scatter plot – summarize
data (identify a typical pattern) using a simple mathematical form.

## Systolic Blood Pressure by Age

```{r,echo=FALSE}
plot(miniH)
abline(lm(BPSysAve ~ Age, data = miniH), col = "red", lwd = 3)
```

## What is the mathematical expression for straight line? {.small}

```{r,echo = FALSE}
plot(BPSysAve ~ Age, data = miniH, type = "n", xlab = "X", ylab = "Y",
    xlim = c(-1, 75), ylim = c(100, 130), main = "Age by SBP")
fit1 <- lm(BPSysAve ~ Age, data = miniH)
abline(fit1)
segments(x0 = 40, x1 = 60, y0 = 117.6449, y1 = 117.6449, lty = "dashed")
segments(x0 = 60, x1 = 60, y0 = 117.6449, y1 = 125.5157, lty = "dashed")
text(50, 115, labels = "20")
text(62, 121, labels = as.character(round(125.5157 - 117.6449, 1)))
text(45, 105, labels = "7.9 (rise) / 20 (run) ~= .359 (B)")
points(x = 0, y = 101.9032, pch = 18, col = "red")
segments(x0 = 1.5, x1 = 5, y0 = 101.9032, y1 = 101.9032)
text(5.5, 101.9032, labels = "Intercept (A)", pos = 4)
text(10, 125, labels = "Y = A + BX", cex = 2)
```

## Estimated regression model {.small}

$\hat{Y} = 101.9 + 0.393 X$

When age is 0, SBP is extrapolated to 101.9.

Slope: 0.393. On average, a difference in SBP by 0.393 mmHg corresponds to a
one-year difference in age

Predicted SBP for someone aged 45:

101.9 + 0.393 x **45** = 119.6

What is the predicted difference in SBP for two persons whose ages are 30 years
apart? (e.g., age 70 vs age 40)

Short way: 0.393 x **30** = 11.79

Long way: (101.9 + 0.393 x **70**) – (101.9 + 0.393 x **40**) =
129.45 - 117.64 = 11.8

