---
title: "Applied Biostatistics 1 Introduction"
subtitle: "Review of Simple Linear Regression"
author: "CUNY Graduate School of Public Health and Health Policy"
output: ioslides_presentation
---

<!--
Welcome to part three of the lecture. I believe students in this class
previously studied the linear regression with one independent variable, which
is called simple linear regression, in an introductory statistics course. So
let’s go through a brief review of simple linear regression.
-->

## Systolic Blood Pressure (SBP) and aging

Hypothesis: age-related physiological changes (narrowing and hardening) of
blood vessels tend to raise blood pressure.

In order to investigate the hypothesis, what would you do first? (With data on
age and blood pressure.)

<!--
Let’s start with an example. Suppose we hypothesize that age-related
physiological changes of blood vessels tend to raise blood pressure, because of
age-related narrowing and hardening of blood vessels. Actually, this may be a
well-established fact rather than a hypothesis. Anyway, the question is, in
order to investigate the hypothesis, what would you do first if you have data
on age and blood pressure?
-->

## Systolic Blood Pressure by Age

* Using NHANES data from 2009-2010 in the `NHANES` R package

```{r,echo=FALSE}
library(NHANES)
data(NHANES)
miniH <- NHANES[NHANES$SurveyYr == "2009_10", c("Age", "BPSysAve")]
miniH <- miniH[miniH$Age >= 18 & complete.cases(miniH), ]
plot(miniH)
```

<!--
In such a case, a typical way to start data analysis is to plot systolic blood
pressure against age, as shown in this figure. Here, the horizontal axis is age
from 18 to 80, and the vertical axis is systolic blood pressure, ranging from
78 to 226. The points are scattered widely. The majority of the points appear
to be in a broad band that goes upward diagonally from left to right,
suggesting a tendency for systolic blood pressure to be higher at older adult
ages. But it is a visual impression I had. Since the points are quite widely
scattered, I am not very confident about my visual impression.
-->

## Local Averaging

* Mean SBP for the 20s, 30s, 40s, ...

```{r,echo=FALSE}
plot(miniH)
lines(
    gplots::wapply(
        miniH$Age, miniH$BPSysAve, method = "width", width = 7, n = 7
    ),
    col = "red", lwd = 3
)
```

<!--
Then the next step may be local averaging. In this figure, the red broken line
was drawn by linking seven points, each of which indicates the mean value of
systolic blood pressure for a 10 year age group. The left end of the red line
indicates that the average systolic blood pressure for those in the 20s is
between 112 and 113, and then the average blood pressure tends to increase with
age up to about 137 for the age group 70 to 80. The red line may appear quite
flat but it is mainly because the vertical axis ranges very widely from 78 to
226, because there are a few outliers whose blood pressure are above 200.
-->

## Linear Regression

We fit a straight line to the points on the scatter plot – summarize
data (identify a typical pattern) using a simple mathematical form.

<!--
Actually, the key concept of simple linear regression is that we fit a straight
line to the points in a scatterplot. By doing so, we summarize the data, or
identify a typical pattern in the data, using a simple mathematical form.
-->

## Systolic Blood Pressure by Age

```{r,echo=FALSE}
plot(miniH)
abline(lm(BPSysAve ~ Age, data = miniH), col = "red", lwd = 3)
```

<!--
This is a straight line that was fitted to the data points using linear
regression. The straight-line goes upward from left to right suggesting that
although the points around the straight line are scattered, the typical pattern
is a rise of systolic blood pressure with age, so this regression result seems
to be consistent with the hypothesis.
-->

## What is the mathematical expression for straight line? {.small}

```{r,echo = FALSE}
plot(BPSysAve ~ Age, data = miniH, type = "n", xlab = "X", ylab = "Y",
    xlim = c(-1, 75), ylim = c(100, 130), main = "Age by SBP")
fit1 <- lm(BPSysAve ~ Age, data = miniH)
abline(fit1)
segments(x0 = 40, x1 = 60, y0 = 117.6449, y1 = 117.6449, lty = "dashed")
segments(x0 = 60, x1 = 60, y0 = 117.6449, y1 = 125.5157, lty = "dashed")
text(50, 115, labels = "20")
text(62, 121, labels = as.character(round(125.5157 - 117.6449, 1)))
text(45, 105, labels = "7.9 (rise) / 20 (run) ~= .395 (B)")
points(x = 0, y = 101.9032, pch = 18, col = "red")
segments(x0 = 1.5, x1 = 5, y0 = 101.9032, y1 = 101.9032)
text(5.5, 101.9032, labels = "Intercept (A)", pos = 4)
text(10, 125, labels = "Y = A + BX", cex = 2)
```

<!--
As you have probably studied in your high school algebra class, the
mathematical expression for a straight line is Y equals A plus B X. For each
point on the straight-line, Y is the value of its location on the vertical
scale, and X is the value on the horizontal axis. 'A' and B are parameters that
define the straight line. B indicates the slope of the straight-line, which
means, on the straight line, if the value of X increases by one, then the value
of Y increases by B.  'A' is called the intercept or the Y intercept, which
means that the diagonal line crosses with the Y axis at 'A'.  In other words,
on the diagonal line, if X equals zero, then Y equals 'A'.
-->

## Estimated regression model {.smaller}

$\hat{Y} = 101.9 + 0.393 X$

When age is 0, SBP is extrapolated to 101.9.

Slope: 0.393. On average, a difference in SBP by 0.393 mmHg corresponds to a
one-year difference in age

Predicted SBP for someone aged 45:

101.9 + 0.393 x **45** = 119.6

What is the predicted difference in SBP for two persons whose ages are 30 years
apart? (e.g., age 70 vs age 40)

Short way: 0.393 x **30** = 11.79

Long way: (101.9 + 0.393 x **70**) – (101.9 + 0.393 x **40**) =
129.45 - 117.64 = 11.8

<!--
The parameters of the straight line equation were estimated for systolic blood
pressure as Y, and age as X, with the N-hanes data. The parameter A was
estimated to be 101 point 9, and the parameter B, the slope, was estimated to
be 0.393. The term on the left hand side is Y-hat, instead of Y, in order to
clearly distinguish observed values of systolic blood pressure, Y, and their
predicted values, Y-hat.  What does 101.9 mean? It may look like systolic blood
pressure of new born babies, because if X equals zero, then Y equals 101 point
9. But the regression model was estimated using data for those aged 18 to 80.
If persons aged under 18 follow the pattern observed for those aged 18 to 80,
101 point 9 may be considered as a predicted systolic blood pressure for age 0.
But the age pattern of blood pressure under age 18 may be quite different from
the pattern for those aged 18 to 80. So, 101 point 9 is considered to be a
hypothetical value of systolic blood pressure at age 0, obtained by
extrapolating the straight line fitted to data for age 18 to 80, to age 0.
0.393 is the slope of the straight line. According to the estimated model, if X
increases by one, then Y-hat increases by 0.393. So it can be considered that
in this data, on average, a difference in systolic blood pressure by 0.393
mmHg, millimeter of mercury, corresponds to a one-year difference in age. “On
average” in this sentence does not exactly mean the arithmetic mean, but means
a representative value, or a typical value, and there is some mathematical
reason for considering the value as representative of different values in the
data.  Let’s play a little with these numbers.  What is the value of systolic
blood pressure for someone aged 45? Suppose someone is aged 45, and we predict
systolic blood pressure of the person based on her or his age only. Then,
according to this regression model, it is 101 point 9 plus 0.393 times 45,
which is 119.6, or about 120.  What is the difference in systolic blood
pressure between two persons whose ages are 30 years apart, like age 70 and age
40?  According to this equation, their typical difference is 0.393 times 30
which equals 15.3 mmHg. Or we can get the predicted values for age 70 and age
40, for example, and then compute their difference.
-->

## Recap

* Simple Linear Regression
* Effect of age on systolic blood pressure using NHANES data

<!--
We reviewed simple linear regression between two varibles, namely systolic
blood pressure and age. We saw that the relationship is fairly linear and we
could use a linear model to explain the relationship.
-->
